---
title: "SEER exploration"
output: html_document
date: "2026-04-12"
---

```{r}
library(tidyverse)
library(janitor)
library(forcats)

# Load Data
raw_df <- read_delim("SEER_brain_2018-22.txt", 
                    delim = "\t", 
                    escape_double = FALSE, 
                    trim_ws = TRUE) %>%
  clean_names() 

glimpse(raw_df)

# printing frequency table of all variables
for (col_name in names(raw_df)) {
  cat("\n--- Variable:", col_name, "---\n")
  print(table(raw_df[[col_name]], useNA = "always"))
}
```

## Cleaning, factoring, and labeling

```{r}
# Comprehensive Cleaning and Standardization of NAs
df_clean <- raw_df %>%
  # Numeric Outliers and Missingness
  mutate(
    # Fix Survival Months (9999 is Unknown)
    survival_months = as.numeric(ifelse(survival_months >= 9999, NA, survival_months)),
    
    # Fix Time to Treatment (999 is Unknown)
    time_from_diagnosis_to_treatment_in_days_recode = na_if(time_from_diagnosis_to_treatment_in_days_recode, 999),
   
     age_factor = factor(age_recode_with_1_year_olds_and_90, 
                        levels = 0:19, 
                        ordered = TRUE) %>%
      fct_recode(
        "00 years" = "0", "01-04" = "1", "05-09" = "2", "10-14" = "3",
        "15-19" = "4", "20-24" = "5", "25-29" = "6", "30-34" = "7",
        "35-39" = "8", "40-44" = "9", "45-49" = "10", "50-54" = "11",
        "55-59" = "12", "60-64" = "13", "65-69" = "14", "70-74" = "15",
        "75-79" = "16", "80-84" = "17", "85-89" = "18", "90+" = "19"
      )) %>%

  
  mutate(
    # Target 9s (Unknown) for specific variables
    across(c(race_recode_white_black_other, radiation_recode, 
             rx_summ_surg_rad_seq, rx_summ_systemic_sur_seq_2007, 
             reason_no_cancer_directed_surgery, diagnostic_confirmation, 
             combined_summary_stage_with_expanded_regional_codes_2004, 
             grade_clinical_2018), ~na_if(., 9)),
    
    # Target 99s and 126s (Unknown/Blank)
    across(c(brain_molecular_markers_2018, median_household_income_inflation_adj_to_2023, 
             rural_urban_continuum_code), ~na_if(., 99)),
    across(c(brain_molecular_markers_2018), ~na_if(., 126))) %>%
  
  # Factor Conversion 
  # Turn NAs into an explicit "Unknown" category for factored variables to avoid selection bias by just removing all NAs (should discuss this with proff)
  mutate(
    grade_factor = factor(grade_clinical_2018) %>%
      fct_recode(
        "WHO Grade 1" = "1",
        "WHO Grade 2" = "2",
        "WHO Grade 3" = "3",
        "WHO Grade 4" = "4",
        "B-Cell/Lymphoma" = "11", 
        "T-Cell/Lymphoma" = "12",
        "NK-Cell/Lymphoma" = "13",
        "Low Grade (L)"   = "15",
        "High Grade (H)"  = "16"
      ) %>%
      fct_na_value_to_level(level = "Unknown"),
    
    molecular_factor = factor(brain_molecular_markers_2018) %>%
      fct_recode(
        "Astrocytoma, IDH-mut (G2)"  = "1",
        "Astrocytoma, IDH-wt (G2)"   = "2",
        "Astrocytoma, IDH-mut (G3)"  = "3",
        "Astrocytoma, IDH-wt (G3)"   = "4",
        "Glioblastoma, IDH-wildtype" = "5",
        "Oligo, IDH-mut/1p19q-codel" = "6",
        "Oligo, G3, IDH-mut/1p19q"   = "7",
        "Medulloblastoma/Other"      = "8",
        "Embryonal/Other"            = "9",
        "Not Applicable"             = "85",
        "Test Ordered/Pending"       = "87"
      ) %>%
      fct_na_value_to_level(level = "Unknown"),
    
    stage_factor = factor(combined_summary_stage_with_expanded_regional_codes_2004) %>%
      fct_recode(
        "Localized" = "1",
        "Regional"  = "2",
        "Distant"   = "7",
        "Benign"    = "8"
      ) %>%
      fct_na_value_to_level(level = "Unknown"),
    
  radiation_factor = factor(radiation_recode) %>%
      fct_recode(
        "None"              = "0",
        "Beam Radiation"    = "1",
        "Radioactive Impl"  = "2",
        "Radioisotopes"     = "5",
        "Refused"           = "7",
        "Recommended, Unk"  = "8"
      ) %>%
      fct_na_value_to_level(level = "Unknown"),
    
    race_factor = factor(race_recode_white_black_other) %>%
          fct_recode(
            "White" = "1",
            "Black" = "2",
            "Other" = "3"
          ) %>%
          fct_na_value_to_level(level = "Unknown"),
    
    # Binary/Simple Factors
    sex_factor = factor(sex) %>%
      fct_recode(
             "Male"   = "1",
             "Female" = "2"
      ),

    # Chemo Labels
    chemo_factor = factor(chemotherapy_recode_yes_no_unk) %>%
      fct_recode(
        "Yes"        = "1",
        "No/Unknown" = "0"
      ),
    surgery_performed = factor(ifelse(reason_no_cancer_directed_surgery == 0, 1, 0)) %>%
      fct_recode(
        "Surgery Performed" = "1",
        "No Surgery"        = "0"
      ) %>%
      fct_na_value_to_level(level = "Unknown"),

    histology_factor = factor(histology_recode_broad_groupings) %>%
      fct_collapse(
        "Germ Cell Neoplasm"       = "25",
        "Neuroepitheliomatous"     = "36",
        "Glioma"                   = "35",
        "Fibromatous Neoplasms"    = "17",
        "Other malignant"          = c("0", "2", "15", "16", "20", "21", 
                                   "28", "30", "32", "34", "38")
      ) %>%
      fct_na_value_to_level(level = "Unknown"),

    primary_site_factor = factor(primary_site_labeled) %>%
      fct_recode(
        "Cerebrum (NOS)"  = "710",
        "Frontal Lobe"    = "711",
        "Temporal Lobe"   = "712",
        "Parietal Lobe"   = "713",
        "Occipital Lobe"  = "714",
        "Ventricle"       = "715",
        "Cerebellum"      = "716",
        "Brain Stem"      = "717",
        "Overlapping"     = "718",
        "Brain (NOS)"     = "719"
      ) %>%
      fct_na_value_to_level(level = "Unknown")
  ) %>%
  
  # Remove old variables
select(-c(site_recode_icd_o_3_who_2008, 
           age_recode_with_1_year_olds_and_90,
            grade_clinical_2018,
            brain_molecular_markers_2018,
            combined_summary_stage_with_expanded_regional_codes_2004,
            radiation_recode,
            race_recode_white_black_other,
            sex,
            chemotherapy_recode_yes_no_unk,
            reason_no_cancer_directed_surgery,
            histology_recode_broad_groupings,
            primary_site_labeled))
```

```{r}
glimpse(df_clean)
for (col_name in names(df_clean)) {
  cat("\n--- Variable:", col_name, "---\n")
  print(table(df_clean[[col_name]], useNA = "always"))
}
```


What kind of question should we assess: 
We could do a prediction on Continuous outcome
How many months of life can a patient expect to gain from surgical intervention, given their specific molecular markers and age?
OR a classification model 
Can we predict the probability of X-year survival based on a patient's molecular profile and socioeconomic status? (Make a binary survival time of Yes or No)

## Classification question 

we need to figure out what a good cutoff would be.
my guess is one year would be a good but will look at sample size changes and distribution of survival outcome between 1 and 2 years.

For 2 year survival time
```{r}
df_2yr_S <- df_clean %>%
  # Drop rows with missing survival outcome
  filter(!is.na(survival_months)) %>%
  
  # Exclude Uncertain cases: 
  # People alive but followed for less than 12 months
  filter(!(vital_status_recode_study_cutoff_used == 1 & survival_months < 24)) %>%
  
  # Create 24-Month Outcome Variable
  mutate(
    survived_24m = factor(ifelse(survival_months >= 24, "Yes", "No"), levels = c("No", "Yes")))

# Check new sample size and balance
nrow(df_2yr_S)
table(df_2yr_S$survived_24m)
```


For 1 year survival time
```{r}
df_1yr_S <- df_clean %>%
  # Drop rows with missing survival outcome
  filter(!is.na(survival_months)) %>%
  
  # Exclude Uncertain cases: 
  # People alive but followed for less than 12 months
  filter(!(vital_status_recode_study_cutoff_used == 1 & survival_months < 12)) %>%
  
  # Create 12-Month Outcome Variable
  mutate(
    survived_12m = factor(ifelse(survival_months >= 12, "Yes", "No"), levels = c("No", "Yes"))
  )

# Check new sample size and balance
nrow(df_1yr_S)
table(df_1yr_S$survived_12m)
```

Since the 12 month is more balanced and keeps more of the sample its probably better


## Looking into amount of surgeries by 12 month survival status
```{r}
# Cross-tab of Surgery vs. 12-Month Survival
table(df_1yr_S$surgery_performed, df_1yr_S$survived_12m, useNA = "always")

prop.table(table(df_1yr_S$surgery_performed, df_1yr_S$survived_12m), 1) * 100
```

Can we predict 1 year survival probability using molecular and SES status
To what extent do molecular biomarkers and SES factors predict 1 year survival in patient with brain cancers?
We could make a model with just the biological factors and compare performance to a model that includes the SES factors as well (median household income, rural vs urban, time to treatment?) and see how performance changes if at all. 


