---
title: "SEER exploration"
output: html_document
date: "2026-04-12"
---

```{r}
library(tidyverse)
library(janitor)
library(forcats)

# Load Data
raw_df <- read_delim("SEER_brain_2018-22.txt", 
                    delim = "\t", 
                    escape_double = FALSE, 
                    trim_ws = TRUE) %>%
  clean_names() 

# printing frequency table of all variables
for (col_name in names(raw_df)) {
  cat("\n--- Variable:", col_name, "---\n")
  print(table(raw_df[[col_name]], useNA = "always"))
}
```

## Cleaning and factoring

```{r}
# Comprehensive Cleaning and Standardization of NAs
df_clean <- raw_df %>%
  # Numeric Outliers and Missingness
  mutate(
    # Fix Survival Months (9999 is Unknown)
    survival_months = as.numeric(ifelse(survival_months >= 9999, NA, survival_months)),
    
    # Fix Time to Treatment (999 is Unknown)
    time_from_diagnosis_to_treatment_in_days_recode = na_if(time_from_diagnosis_to_treatment_in_days_recode, 999),
    
    # Convert Age to Continuous using the confirmed 5-year midpoint math
    # 19 was confirmed as 90+, so this creates a linear scale
    age_continuous = (age_recode_with_1_year_olds_and_90 * 5) + 2.5
  ) %>%
  
  # Filter Out Missing Outcome (still need to determine this)
  # filter(!is.na(survival_months)) %>%
  
  # Targeted NA Standardization for Categorical Codes
  mutate(
    # Target 9s (Unknown) for specific variables
    across(c(race_recode_white_black_other, radiation_recode, 
             rx_summ_surg_rad_seq, rx_summ_systemic_sur_seq_2007, 
             reason_no_cancer_directed_surgery, diagnostic_confirmation, 
             combined_summary_stage_with_expanded_regional_codes_2004, 
             grade_clinical_2018), ~na_if(., 9)),
    
    # Target 99s and 126s (Unknown/Blank)
    across(c(brain_molecular_markers_2018, median_household_income_inflation_adj_to_2023, 
             rural_urban_continuum_code), ~na_if(., 99)),
    across(c(brain_molecular_markers_2018), ~na_if(., 126)),
    
    # Target 7s and 8s (Ambiguous treatment status)
    across(c(radiation_recode, reason_no_cancer_directed_surgery), 
           ~ifelse(. %in% c(7, 8), NA, .)),
    
    # Targeted sweep for Molecular Markers (85=N/A, 87=Not documented)
    brain_molecular_markers_2018 = ifelse(brain_molecular_markers_2018 %in% c(85, 87), NA, brain_molecular_markers_2018)
  ) %>%
  
  # Factor Conversion 
  # Turn NAs into an explicit "Unknown" category for factored variables to avoid selection bias by just removing all NAs (should discuss this with proff)
  mutate(
    grade_factor = fct_na_value_to_level(factor(grade_clinical_2018), level = "Unknown"),
    molecular_factor = fct_na_value_to_level(factor(brain_molecular_markers_2018), level = "Unknown"),
    stage_factor = fct_na_value_to_level(factor(combined_summary_stage_with_expanded_regional_codes_2004), level = "Unknown"),
    radiation_factor = fct_na_value_to_level(factor(radiation_recode), level = "Unknown"),
    race_factor = fct_na_value_to_level(factor(race_recode_white_black_other), level = "Unknown"),
    
    # Binary/Simple Factors
    sex_factor = factor(sex),
    chemo_factor = factor(chemotherapy_recode_yes_no_unk),
    histology_factor = fct_na_value_to_level(factor(histology_recode_broad_groupings), level = "Unknown"),
    primary_site_factor = fct_na_value_to_level(factor(primary_site_labeled), level = "Unknown"),
    surgery_performed = fct_na_value_to_level(
      factor(ifelse(reason_no_cancer_directed_surgery == 0, 1, 0)), 
      level = "Unknown"
    )
  ) %>%
  # Remove the constant columns (brain cancer ICD code)
  select(-c(site_recode_icd_o_3_who_2008))

# Verification
glimpse(df_clean)
colSums(is.na(df_clean))

for (col_name in names(df_clean)) {
  cat("\n--- Variable:", col_name, "---\n")
  print(table(df_clean[[col_name]], useNA = "always"))
}
```

What kind of question should we assess: 
We could do a prediction on Continuous outcome
How many months of life can a patient expect to gain from surgical intervention, given their specific molecular markers and age?
OR a classification model 
Can we predict the probability of X-year survival based on a patient's molecular profile and socioeconomic status? (Make a binary survival time of Yes or No)

## Classification question 

we need to figure out what a good cutoff would be.
my guess is one year would be a good but will look at sample size changes and distribution of survival outcome between 1 and 2 years.

For 2 year survival time
```{r}
df_2yr_S <- df_clean %>%
  # Drop rows with missing survival outcome
  filter(!is.na(survival_months)) %>%
  
  # Exclude Uncertain cases: 
  # People alive but followed for less than 12 months
  filter(!(vital_status_recode_study_cutoff_used == 1 & survival_months < 24)) %>%
  
  # Create 24-Month Outcome Variable
  mutate(
    survived_24m = factor(ifelse(survival_months >= 24, "Yes", "No"), levels = c("No", "Yes")))

# Check new sample size and balance
nrow(df_2yr_S)
table(df_2yr_S$survived_24m)
```


For 1 year survival time
```{r}
df_1yr_S <- df_clean %>%
  # Drop rows with missing survival outcome
  filter(!is.na(survival_months)) %>%
  
  # Exclude Uncertain cases: 
  # People alive but followed for less than 12 months
  filter(!(vital_status_recode_study_cutoff_used == 1 & survival_months < 12)) %>%
  
  # Create 12-Month Outcome Variable
  mutate(
    survived_12m = factor(ifelse(survival_months >= 12, "Yes", "No"), levels = c("No", "Yes"))
  )

# Check new sample size and balance
nrow(df_1yr_S)
table(df_1yr_S$survived_12m)
```

Since the 12 month is more balanced and keeps more of the sample its probably better


## Looking into amount of surgeries by 12 month survival status
```{r}
# Cross-tab of Surgery vs. 12-Month Survival
table(df_1yr_S$surgery_performed, df_1yr_S$survived_12m, useNA = "always")

prop.table(table(df_1yr_S$surgery_performed, df_1yr_S$survived_12m), 1) * 100
```

